\documentclass[a4paper,10pt,twocolumn,twoside]{article}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{filecontents}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{listings}
\usepackage{authblk}
\usepackage{caption}
\usepackage{mathtools}
\usepackage{underscore}
\usepackage{url}
\usepackage{rotating}

\usetikzlibrary{patterns,shapes,positioning,calc}

\lstset{language=C, basicstyle=\ttfamily\small, columns=flexible}

\pgfplotsset{compat=1.8}

\author{Lars Kirkholt Melhus}
\author{Rune Erlend Jensen}
\affil{Norwegian University of Science and Technology}
\title{Measurement Bias from Address Aliasing}
\date{} % This is a timeless piece

% Pretty printing of performance counter names
\newcommand{\perfctr}[1] {
  {\lowercase{#1}}
}

\begin{document}

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract} % Should be 150 -- 200 words max
    % Motivation -- Problem statement -- Approach -- Results -- Conclusions

    In this paper we identify the underlying mechanics resulting in measurement bias on modern Intel microarchitectures. 
    The magnitude of this performance bias has been shown to be significant, and can be as high as a factor two.
    
    We analyze a small program that was presented as a curiosity in related previous work on \emph{measurement bias};
    Code that for unexplained reasons performs significantly worse than average for certain configurations of system environment variables.

    Using hardware performance counters, we identify a strong correlation with resource stalls, ultimately caused by address aliasing events.
    The CPU uses an optimistic strategy when issuing operations out of order.
    Only the last 12 address bits are compared to determine if there was a conflict, sometimes resulting in performance penalties from false dependencies.

    We show how address aliasing can cause measurement bias for two particular external properties; environment variable size, and dynamically linked heap allocation libraries.
    Performance penalties from unfavorable memory address allocations can be more than double the cycle count, even for real world applications.
    Furthermore, the dynamic memory allocation pattern for common libraries like libc tends to give worst case performance by default for large allocations.

    \end{abstract}
    \bigskip
        
  \end{@twocolumnfalse}
]


\section{Introduction}
% - Measurement bias in systems research
%   - What is measurement bias
%     - fundamentally anything external
%     - Room temperature
%   - Studied before
%     - External properties that affect memory layout
%       - link order (not really)
%       - env size
%   - Has been shown to be significant and commonplace
%     - But difficult to deal with
%     - Need to apply statistical methods to experiments
%       - randomization of experimental setups
% - Want to look at flipside: understand what why bias occurs
%   - Use for optimization
%   - Find interesting cases
%   - Known causes: cache alignment, ignore.
% - Analysis: 
%   - Use known example that exhibit bias
%   - Find lesser known source in address aliasing
% - Results:
%   - Applicable general knowledge about Intel architecture
%   - Useful for optimization
%   - Libraries and compilers fail to account for it in reality.

How fast does an algorithm run?
Is program A faster than program B on \emph{my} machine?
These types of questions are, perhaps surprisingly, not always trivial to answer.
The usual approach to quantify performance is to simply use some metric, such as the wall time or number of cycles executed.
However, performance is not just a function of the collection of instructions in an executable, but also dependent on various \emph{external} factors.

In other fields of science, accounting for sampling bias is well understood.
Trying to determine the political opinion of a country just by sampling a narrow geographical area will probably not be very accurate.
The same concept applies for software; measuring once is analogous to sampling only a tiny geographical area in the set of possible system configurations.

There are many external variables that has the potential to affect system performance.
In particular, anything which can offset memory addresses of code or data at runtime can be significant.
Memory accesses interacts with low level hardware buffers, which can heavily favor certain memory layouts over others.
Consequentially, any environmental factor that can alter memory layout has a potential performance impact.
Failing to account for these effects lead to \emph{measurement bias}. 

% This is basically former background section
Measurement bias has been shown to be commonplace and significant in real applications~\cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
Previous work looks at environmental factors such as the size of Unix environment variables, which affects memory by offsetting the initial position of stack.
Because of the complexity of modern hardware, it is often impossible to predict exactly how seemingly innocuous things such as environment variables ultimately affect program execution.
While effects of bias has been shown to be commonplace and significant, they are unpredictable and difficult to deal with.
Without an accurate understanding of the underlying memory interactions, the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts.
Previous work focuses mostly of how to account for bias by proper sampling, using randomization of execution contexts and statistical methods. % Cite is basically the same as above, all three are about the same -- difficult to split. 

Our approach is instead to try to understand some of the underlying reasons to why bias can occur, in an attempt to utilize this knowledge for potential optimization. % Could have a sentence with cite of Blind Optimization.
Usually, the explanation will be that different memory layouts interact with various hardware buffers.
Certain mechanisms such as cache contention or branch misprediction are already well understood, thus we are looking for the more esoteric processor features.
Analyzing a program that exhibits strong bias from environment size, collecting about 200 events using hardware performance counters, we find strong correlation with resource stalls from \emph{address aliasing}.
Aliasing is an artifact of a much lesser known, presumably Intel specific, optimization on the out-of-order execution pipeline.
In short, the CPU uses a heuristic for determining whether memory loads are dependent on previous writes, comparing only the last 12 virtual address bits.
False dependencies incur performance penalties, and we show how this can explain many common cases of measurement bias.

In addition to environment variables, we look at dynamic heap allocation libraries as an external factor that can trigger (or amplify) bias resulting from address aliasing.
We find that most libraries tend to produce worst case behavior by default with respect to aliasing, caused by favoring page alignment for large allocations.
The cost of address aliasing can be significant, exemplified with a simple program yielding more than 50~\% speedup based on heap address alignment alone.
% [This next part might be added to the abstract in some form:] 
Fortunately, with an understanding of the underlying mechanisms, measurement bias from address aliasing can often be quite accurately predicted.
For some applications it might even make sense for programmers to actively take aliasing into account for architecture specific optimization, as neither compilers nor heap allocators seem to handle it well.

% Include preview of aliasing program early to provike interest. 
% Also lightly referenced in introduction/background
\begin{lstlisting}[float=t, language=C, caption={Micro-kernel presented by Mytkowitz et al~\cite{Mytkowicz:2009:WrongData}, succeptible to bias from variations to system environment variables.}, label={lst:loopkernel}, frame=lines]
static int i, j, k;
int main() {
    int g = 0, inc = 1;
    for (; g < 65536; g++) {
        i += inc;
        j += inc;
        k += inc; 
    }
    return 0;
}
\end{lstlisting}



\subsection{Methodology}
% - Goal: Produce empirical data showing what causes bias
% - Two pitfalls:
%  - Observer effect
%  - Measurement bias
% - Hardware performance counters
%   - MSR, architecture specific
%   - Closest thing to the CPU
%   - Use perf-stat
%   - Control cache metrics
% - Best practices for controlled experimental setup
%   - No ASLR, what it is
%   - No hyperthreading, disable frequency scaling.
% - Experimental setup
%  - Intel haswell architecture
%    - Restrict to a single cpu to limit scope

In order to understand bias behavior it is important to be able to do precise measurements.
Besides measurement bias in itself, one must also consider \emph{observer effect}.
Any kind of instrumentation added to the code, such as a counter variable for each invocation of a function, has the potential to produce misleading results~\cite{Mytkowicz:2008:OE&MB}.
Instead, we use instrumentation provided directly by the hardware; in the form of performance counters. 
Modern processors have a set of model specific registers (MSRs) that can be configured to \emph{count} various events, such as cycles executed, branch misses, instructions fetched, and so on.
Recent Intel architectures have several hundred different available events, which can give a very detailed view of what happens inside the CPU.

Performance counters are supported in the Linux kernel, and we access them via a tool called \emph{perf}. 
We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the Intel reference manual \cite{Intel:2013:Volume3B}.
A small Python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
The same script simulates changes to environment size, which is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{Intel:2012:OptimizationManual}.
Results are averaged over multiple runs to reduce potential random error, supplying the \texttt{-r} option to perf-stat. % Not median, good question. man perf-stat says average + std.dev. 

Because bias can be hardware dependent, and to keep the scope of this work manageable, we choose to focus on the Intel ``Haswell'' microarchitecture specifically.
Our setup consists of a 4th generation Intel\textsuperscript{\textregistered{}} Core\texttrademark{} i7-4770K processor, running 64-bit Ubuntu 14.04 LTS on kernel version 3.13.0-24-generic.
Code samples are compiled using the {\small GCC} toolchain, version 4.8.2-19ubuntu1.

We also have to ensure we are not affected by measurement bias beyond what we are actually trying to observe, and in general follow best practices on gathering data~\cite{Mytkowicz:2009:WrongData}.
Most importantly, this means keeping the memory address space under control.
For security reasons, addresses of stack, heap and dynamic libraries are often randomized at load time, a technique known as \emph{Address Space Layout Randomization}~\cite{Shackham:2004:ASLR}. % Should maybe have more generic reference
By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
All benchmarks are done on a machine under minimal load, and \emph{frequency scaling} is disabled to keep the CPU's clock speed fixed.
Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention between threads.



% === Random notes section on 4K alias ===
% There seems to be two separate issues here
% 1: store-to-load forwarding speculative, then discover that the addresses are not the same. (pp 500)
% 2: store-load issuing the load first speculatively. WAR conflict in MOB, reissue load. VTune event "4K alias"
%
% MOB: "Accepts and processes all memory operations". pp 62
%      Allows issue load before preceding store, speculatively


% Discussed in Optimization manual (pp. 144, 403, 564, 583)
% - Said to mean many things
%   - "False store forwarding" -- Only checks last 12 bits. Does it attempt to forward?) (pp. 564)
%   - Loads that have partial address match with preceding stores, causing the load to be reissued. 
%     Estimate 5 cycle cost
%   - "False dependencies in MOB due to partial compare on address"
%   - Very clear separate event in VTune Amplifier documentation online 
% - Detected by LOAD_BLOCK.OVERLAP_STORE on previous architectures (pp. 599)
% - Present event code and description
% Optimization guidelines
% - Has some cost associated with it
% - Intel suggests measures for mitigating cost
%   - Change offsets between input and output buffers
% Potentially bias inducing
% - Anything that affects memory layout
%   - Stack variable offset from env size
%   - Heap addresses from dynamic allocation libraries

% User/Source Coding Rule 8. (H impact, ML generality) Consider using a special memory allocation
% library with address offset capability to avoid aliasing. .................................................. 3-55
%   > Examples mention avoiding L1 and L2 cache conflicts.
% User/Source Coding Rule 9. (M impact, M generality) When padding variable declarations to avoid
% aliasing, the greatest benefit comes from avoiding aliasing on second-level cache lines, suggesting
% an offset of 128 bytes or more ................................................................................... 3-55

% Facts
% * Loads can be issued before stores, speculating the store is not going to be to a conflicting address. (pp. 62)
% * All memory operations go through the memory order buffer. (pp. 63)
% * Memory disambiguation has to do with prediction. Predicted ok load from L1. (pp. 65)
% * Example of speculative execution of loads, with mem. disambiguation (pp 131)
% * Rule 56, 8 & 9. Aliasing case wrt Store forwarding. Claims 4K causes stall (not reissue) (pp. 141)
% * Statement: 4K aliasing causes stall = included in LOAD_BLOCK.OVERLAP_STORE
% * For Sandy Bridge, description of LD_BLOCKS_PARTIAL.ADDRESS_ALIAS states reissue. (pp. 583)



\section{4K Address Aliasing}
% Mechanisms for increased throughput and parallelism of memory operations
% In particular loads that might depend on earlier stores
% - Worst case have to wait for store to retire
% - Supports /speculative/ issue based on prediction on whether load conflicts with previous store.
%   - Later verified (memory disambiguation)
% - If refer to same address, take value from sotre buffers via /store forwarding/. 
%   - Don't have to wait for store to finish
% An event known as 4K aliasing occurs when the load and store addresses match in the lower 12 bits
% - CPU detects /false/ dependencies between memory operations.
% - Counted by event ...
% - Sometimes refered to as /false store forwarding/. (pp. 564)
% Optimization guidelines
%   - Avoid store and non-dependent load with 4K between (rule 56)
%   - Special memory allocator (rule 8) -- for 64 KiB alias
%   - Padding variables (rule 9) -- For 64 KiB alias
% - All classified as "High" impact.
% - We demonstrate how this effect causes bias for memory allocator and store/load single
% - How programmers can adjust layout manually
Modern processors are \emph{superscalar}, and achieves parallelism by issuing multiple instructions simultaneously and out of order.
One of the issues that can limit throughput is dependencies between a load and previous stores.
Typical software consists of about 38~\% memory operations, with about two thirds of them being loads~\cite{Intel:2006:InsideICM:SmartMemoryAccess}.
More often than not, loads can safely be issued before a previous store has completed and written its value to L1 cache.
Loads can therefore be issued \emph{speculatively}, based on a prediction on whether it will conflict with a previous store.
If the load and store locations are the same, the value can be \emph{forwarded} from the store before it retires.

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096.
A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
The number of times this happens can be counted by the following performance counter:
\begin{description}
  \item{LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS}\marginpar{\vspace{0.0pt}\footnotesize\begin{tabular}{@{} l @{\hskip 2pt} l }Event & Mask \\ 0x07 & 0x01\end{tabular}} \hfill \\
  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.~\cite{Intel:2012:OptimizationManual}
\end{description}

As address aliasing depends on the memory addresses of loads and stores, environmental factors that affects memory has the potential to induce aliasing conditions.
In the following sections, we show how performance penalties from address aliasing can be the root cause of measurement bias.



\section{Bias from Environment Size}
% General intro on why environment triggers aliasing
% - Image of memory organization, this can be generic.
%   - Each area affected by randomization differently
% - Default 16 bit stack alignment guarantee.
% Analysis section
% - Present data on correlation
% - Plot graph of 2 x 4K areas. 
% - Addresses of static variables from symbol table
% - Addresses of automatic variables from asm output
% - Outline different scenarios that can happen for aliasing with 1 or 2 variables
% - Conclusion: was bias from false dependencies
% Avoid aliasing for env size
% - Show altered program
% - Proof of concept, in theory possible for compilers to generate such code.

Unless the program explicitly accesses environment variables, it is not the environment variables themselves that are important, but rather the effect their total size has on the alignment of stack.
Environment variables and program arguments are allocated in the stack section of virtual memory, before the first call frame.
Changing environment variables will therefore offset the addresses of stack, and consequently all stack allocated variables.

% Generic intro to memory layout (with focus on stack/env interaction)
Figure \ref{fig:virtualmemory} shows the relative positions of some important sections of memory at runtime, assuming a 64-bit process mapped to virtual memory. 
The stack is normally placed close to the upper address 0x7fff'ffffffff\footnote{Modern processors do not actually use the full 64-bit space, only the low order 47 bits are used for addressing memory}, but offset by environment variables.
There is still some level of stack alignment happening, independent of environment offsets. 
Normally, the stack is aligned to 16 bytes, which will be enforced by the compiler. % [ref]
Within a span of 4096 bytes there are thus 256 possible initial addresses, each representing a different execution context.
Any one of these contexts could trigger aliasing issues between stack and some other area of memory remaining unaffected by environment size.
This includes dynamic memory allocateded on heap, memory mapped files, as well as statically allocated data. 

Note that there is no clear relationship between environment size and stack location with ASLR enabled.
However, there will still be as many execution contexts with respect to aliasing (considering stack only), making any occurrences of measurement bias indeed random.
% In addition to affecting stack, ASLR also randomizes location of dynamically linked files (memory mapped), and initial position of heap.
% Moving to the lower end of the address space, we find the text, data and bss sections, which contains program code and statically allocated data.
% The memory addresses of these sections are at compile time by the linker, and can be determined by inspecting the executable.

\begin{figure}[float=t]
  \caption{Organization of virtual memory for a 64-bit process.}
  \label{fig:virtualmemory}
  \begin{tikzpicture}[font=\footnotesize]

    % See page 453 in pgf manual
    \node [
      rectangle split, rectangle split parts=10, 
      rectangle split part fill={white, white, lightgray, white, lightgray, white, lightgray, white},
      draw, anchor=center, text width=2cm
    ] (m)
      {
        \nodepart{one}
          environment
        \nodepart{two}
          stack
        \nodepart{four}
          mmap area
        \nodepart{six}
          heap
        \nodepart{eight}
          bss
        \nodepart{nine}
          data
        \nodepart{ten}
          text
      };

    \draw [decorate, decoration={brace, amplitude=5pt}] (m.south west) -- (m.seven split west) 
      node [black, midway, xshift=-1.5cm, text width=2.0cm] 
        { Program code and static data } ;

    \node[right] at (m.north east)
      { \scriptsize{0x7fff'ffffffff} } ;

    \node[right] at (m.south east)
      { \scriptsize{0x400000} } ;

  \end{tikzpicture}
\end{figure}

\subsection{Microkernel Analysis}
To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by T. Mytkowicz et. al.~\cite{Mytkowicz:2009:WrongData}, reproduced here in Listing~\ref{lst:loopkernel}. 
This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the example code is simple and straight forward to analyze.
Still, no satisfactory explanation as to what actually causes bias was given in the original paper.

% Large plot of cycle count for two 4K stack sections.
\pgfplotstableread{bin/micro-kernel-cycles.dat}{\stackoffsettable}
\begin{figure*}[t]
  \caption{Bias from environment size for micro-kernel in Listing~\ref{lst:loopkernel}. Measured average of 100 cycle count samples for 512 different environments. Spikes show aliasing case, occurring once for each 4K period}
  \label{fig:envbias}
  \begin{tikzpicture}
    \begin{axis}[
        width = \textwidth,
        height = 6cm,
        font = \small,
        xlabel=Bytes added to environment,
        ylabel=Cycles,
        domain = 0:8192,
        xtick = {0,1024,...,8192},
        xmin = 0,
        xmax = 8192,
        cycle list name = exotic
      ]
      \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \stackoffsettable ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

Performance counter measurements of cycle counts are shown for 512 different environment sizes in Figure~\ref{fig:envbias}.
We measure every 16 byte increment of environment size, covering two 4K periods of initial stack addresses.
A finer sampling is not necessary, because the stack is by default aligned to 16 byte.
The program is compiled with {\small{GCC}} using no optimization, as any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.

There are clearly two worst cases, indicated by significant spikes at the 3200 and 7296 byte offsets.
We sampled an extensive set of performance counters in addition to cycle count for each execution context.
By looking at the \emph{median} value of each counter compared to the extreme cases, we narrowed down the performance events that behaved similarly to cycle count.
Table \ref{tab:loopcorrelation} shows numerical counter values for median over all environments, compared to worst cases of 3200 and 7296 bytes added.

% Detailed comparison between median and worst cases.
\begin{table}[t]
  \caption{Events with significant correlation to cycle count.}
  \label{tab:loopcorrelation}
  \centering
  \renewcommand{\tabcolsep}{1pt} % Less padding between cells
  \pgfplotstabletypeset[
    font=\scriptsize,
    int detect, % Output whole numbers for counter values
    col sep=comma,
    columns={Performance counter, Median, [index]3, [index]4},
    column type=r,
    columns/Performance counter/.style={
      string type, 
      column type=l,
      postproc cell content/.append code={
        \pgfkeysalso{@cell content=\perfctr{##1}}
      }
    },
    every head row/.style={
      output empty row,
      before row={\toprule
        Performance counter & Median & Spike 1 & Spike 2 \\
        },
      after row=\midrule,
    },
    every last row/.style={after row=\bottomrule}
  ]{bin/micro-kernel-comparison.csv}
\end{table}

Notice the most extreme change from median to worst case is clearly the number of alias events.
If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
Some of the other counters are less interesting; for example, bus- and cpl cycles, which will naturally vary with the total cycle count.
These counters are obviously not indicative of any causal relationship, and are omitted.

% Need input here, does this make sense?
On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from about 270,000 to about 135,000.
The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Intel:2013:Volume3B}.
Fewer stalls in the aliasing case could indicate less contention on the reservation station.
This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports seem to have fewer operations going in parallel.
Note that the number of micro-ops \emph{retired} overall does not change, but for some reason the micro-ops stays in the execution pipeline for a shorter time.
Our assumption is that this is a by-product of stalls generated from aliasing.

% More deep dive section (maybe separate headline...)
\subsection{The Offending Variables}
These numbers clearly points to address aliasing as a plausible explanation, thus the next step is to see exactly which memory accesses are aliasing. 
For that we need to know the addresses of all variables at runtime.
The program contains five variables; \texttt{g} and \texttt{inc} which are stack allocated, and \texttt{i}, \texttt{j} and \texttt{k} which are statically allocated.
Virtual memory layout of static data is decided at compile time, and we can find the location of these variables by looking at the symbol table in the ELF executable.
We find the addresses to be \texttt{\&i} = 0x60103c, \texttt{\&j} = 0x601040, and \texttt{\&k} = 0x601044.\footnote{ELF symbol tables can be read using \texttt{readelf -s}}

Observing addresses of stack allocated data at runtime is more challenging, as we have to make sure to not introduce any observer effects that alters the addresses as we are observing them.
A small amount of assembly code was added to calculate the addresses of \texttt{g} and \texttt{inc}, outputting to stdout directly using the \texttt{syscall} instruction.
Making no change to the stack allocation instructions and not affecting the addresses of static variables, the instrumented program have the exact same bias to environment size, free from observer effects.
We find that the spike in cycle count occurs precisely when the address of \texttt{inc} alias with \texttt{i}, the first spike happening when \texttt{g} is at 0x7fffffffe{\underline{038} and \texttt{inc} is at 0x7fffffffe{\underline{03c}}.
The assembly output of gcc for the inner loop is shown below, indicating the conflicting load and store instructions.

% Refer to relevant section of the generated assembly file (in gas syntax)
\lstinputlisting[
  linerange={22-38}, 
  numbers=left,
  firstnumber=22
  ]{bin/micro-kernel-annotated.s}

Because the stack is aligned to 16 bytes, or 4 words, there are a couple of different scenarios that could have been observed here.
Static variables are fixed and covers 12 contiguous bytes (3 words), in our case the addresses end in 0xc, 0x0 and 0x4, leaving the 0x8 slot free.
Automatic variables occupies 8 contiguous bytes on stack (2 words), in our case the addresses will always fit in the 0x8 and 0xc slots..
In this scenario, \texttt{g} will never alias with any of the static variables -- as it always covers the 0x8 slot not occupied by either of \texttt{i}, \texttt{j} or \texttt{k}.
A less fortunate scenario with respect to the number of alias events occurs when there can be collisions with both stack allocated variables, which can be achieved for example by reserving an extra 8 bytes to offset \texttt{i}, \texttt{j} into the 0x8, 0xc slots. 
While this will give significantly more alias counts, it has little effect on the number of cycles executed.

In conclusion, we identified that address aliasing is the root cause of measurement bias from environment size for this program.
Worst case occurs for precisely one out of 256 possible initial stack addresses in every 4K segment, where resource stalls are generated because of false dependencies between stack and static data.
The program is \emph{biased} towards environment sizes that avoids this specific stack alignment.


\subsection{Avoiding Aliasing}
Addresses of automatic variables can not be determined statically, because the position of stack at runtime is generally unknown. 
In addition to being offset by environment variables, the stack address can also be perturbed by other factors such as ASLR. 
Although one can not easily know if a collision is going to happen for a given environment, it is possible to change the program to account for possible alias effects.
Shown in Listing \ref{lst:loopfixed} is a proof of concept of how alias-free code could be generated in this particular case.
If the addresses do alias, a branch to an alternative but semantically equivalent code path is performed.
Calling the function recursively will effectively allocate a new set of variables a bit further down the stack, and the alias condition is avoided.

% See analysis for results showing this does not alias
\begin{lstlisting}[float=t, language=C, caption={Dynamically detect aliasing case, and avoid by pushing another stack frame.}, label={lst:loopfixed}, frame=lines]
#define ALIAS(a, b) \
    ((((long)&a)&0xfff)==(((long)&b)&0xfff))
static int i, j, k;
int main() {
    int g = 0, inc = 1;
    if (ALIAS(inc, i) || ALIAS(g, i))
        return main();
    for (; g < 65536; g++) {
        i += inc;
        j += inc;
        k += inc;
    }
    return 0;
}
\end{lstlisting}

While this particular solution is clearly not something one would want to do in practice, it at least shows that compilers and programmers \emph{could} take measures to account for aliasing.
This becomes much more enticing when the performance penalties are more significant, which we find is often the case for aliasing heap allocations.



\section{Bias from Heap Allocation}
% Intro
% - Bias from aliasing caused by accesses to any part of memory
% - Dynamic memory usually from dynamically linked libs = bias
% - More interesting: default behavior provokes aliasing 
% - Intel warning, copy routine. 
% - Trouble with aligned read/write buffers, creating constant conflicts in sliding window.
% Heap allocation alias by default
% - Normal heap with brk/sbrk syscalls
%   - Manage heap and increase on demand
% - Different depending on size (and other parameters)
%   - Use mmap for performance, usually large allocations
%   - Posix function, anonymous memory mapping
%   - Page aligned
% - Normal addr. vs mmap addr.
% - Makes no difference ++ everything is conceptually the same heap, but alias by default!
% Convolution kernel Analysis
% - More contrived example; convolution
%   - Simple code disregarding endpoints
%   - Manipulate start addresses by offseting pointers
%   - Compiled with gcc O3, shows bad performance (graph) + discussion
% - Table with lots of data points, showing actual numbers for each offset
% Architectural optimization
% - Modified example with restrict and -march=native
% - Vectorization -- probably less loads that causes reduction mostly
% Bahavior of other Allocators
% - Intro each allocator
% - Present table
% Conclude
% - not good...
% - Not tested other compilers, not the purpose of this paper. 
% - How did we find this example? Intelligent design, literally first guess

Address aliasing can be caused by conflicting pairs of load and store operations to any part of memory.
In the previous sections we saw collisions between static data and stack, observing bias from external conditions that affected addresses of automatic variables.
Most dynamic memory is allocated on the \emph{heap}, which is managed by an \emph{allocator}.
A heap allocator is responsible for managing dynamic memory, and ultimately assigns the actual addresses of heap allocated variables at runtime.
Heap allocation routines such as \texttt{malloc} and \texttt{free} are typically dynamically linked, for example as part of glibc, or some other library.
The particular library used therefore constitutes an important part of the execution context, as linking to a different library, or a library with some alternative configuration, can impact heap addresses at runtime.


\subsection{\texttt{malloc} will Alias by Default}
Acquiring dynamic memory at runtime is usually done by calling \texttt{malloc}, which takes a number of bytes to allocate as input, and returns a pointer to that area.
A typical malloc implementation uses different strategies for how to allocate memory depending on the \emph{size} of the request.
\begin{itemize}
  \item The ``regular'' heap area in virtual memory is used for smaller allocations. 
  Programs are initially given a relatively small heap, which the allocator can increase by calling \texttt{brk} or \texttt{sbrk} system calls.

  \item For larger allocations, it can be more efficient to use anonymous memory mapping by calling the Posix function \texttt{mmap} instead of growing the regular heap.
  One drawback of this is that it requires the memory to be zero initialized, and whether memory mapping is used or not is based on some cost heuristic employed by the allocator.
\end{itemize}
The heap section starts at a relatively low address (close to the data section), growing upwards.
Memory mapped chunks are placed towards the upper end of the virtual address space, closer to the stack.
Whether a memory allocation request is served by the heap or by memory mapping is therefore easy to determine just by looking at the actual pointer values returned.
While addresses in the regular heap can look something like 0x16e30a0 or 0x1723020, pointers returned by \texttt{mmap} are numerically much larger -- for example 0x7f0318a8f010 or 0x7f03105d2010.
This distinction is of course unimportant for application developers, as everything is conceptually the same ``heap''.
However, \texttt{mmap} has an interesting property in that allocations will always be page aligned.
The page size is 4096 bytes, meaning two pointers returned by \texttt{mmap} will \emph{always} alias.\footnote{libc's version of malloc adds 16 bytes of metadata at the beginning, therefore every memory mapped address end with 0x010.}
This behavior is often the worst case for functions that operate on two or more independent buffers, as we will see in the following sections.

\subsection{Aligned Sequential Access}
Many functions operate in a ``sliding window'' fashion; reading and writing to different buffers in some loop construction.
This type of program is potentially vulnerable to 4K aliasing, where the worst case will be when the read and write pointer addresses are aliased, continuously generating false conflicts.
As an example of this, consider a naive implementation of linear correlation shown in Listing \ref{lst:conv} (for simplicity, endpoints are not handled).
The function containins lots of interleaved loads and stores, iterating over two independent memory buffers in a tight loop.
We will see that the performance of this program greatly depends on the address alignment of each buffer, favoring memory addresses that are not closely aligned on the last 12-bits.

\lstinputlisting[
  float=t, 
  language=C, 
  caption={Naive implementation of convolution. Highly sensitive to aliasing between input and output arrays.},
  label={lst:conv},
  frame=lines
]{bin/convolution-kernel.c}

% Use mmap directly instead maybe?
By using an input size of $n=2^{20}$ (4 MiB in memory for each array), glibc's heap allocator ends up always choosing \texttt{mmap} to serve requests.
By default, even with address randomization enabled, both pointers will have the same address suffix of 0x010.
To analyse performance for different addresses of input and output buffers, we manually insert padding to offset one of the memory mapped pointers.
This is accomplished by requesting a bit more memory, and use pointer arithmetic to offset the input arguments when calling \texttt{convolution}.
Controlling the offset parameter, we can create environments where the inputs are some number of \texttt{sizeof(float)} bytes apart.
Allocating and managing these buffers at startup takes a non-neglegible amount of work.
The overhead can be masked by repeatedly invoking the \texttt{convolution} kernel after allocating and initializing all the inputs.
\begin{lstlisting}
    for (i = 0; i < k; ++i)
        convolution(n, input, output + offset);
\end{lstlisting}
An estimate of the actual cost of one invocation can be calculated by averaging over a number of repeated function calls, after subtracting the constant overhead of invoking it only once.

\[
t_{\text{estimate}} = \frac{t_{k} - t_{1}}{k - 1}
\]

Our results are from using $k=11$, effectively using the average over 10 loop iterations to estimate the cost of a single invocation.
For every run, performance counter measurements is also averaged over 10 samples, using the repeat mechanism of \texttt{perf-stat}.

\pgfplotstableread{bin/conv-default-o2.estimate.dat}{\convtabletwo}
\pgfplotstableread{bin/conv-default-o3.estimate.dat}{\convtablethree}
\begin{figure*}[t]
  \caption{Estimated cycle- and alias counts for different offsets between input and output arrays in convolution kernel from Listing~\ref{lst:conv}. Offset 0 means equal 12-bit address suffix, which is default behavior for \texttt{mmap} allocations. Input size $n=2^{20}$.}
  \label{fig:conv-default}
  \begin{tikzpicture}
    \begin{axis}[
        title=\texttt{cc -O2},
        font = \small,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtabletwo ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtabletwo ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        title=\texttt{cc -O3},
        font = \small,
        xlabel=Relative offset in \lstinline!sizeof(float)! bytes,
        cycle list name=black white
      ]
      \addplot table[x expr = \thisrowno{0}, y = cycles:u] \convtablethree ;
      \addplot table[x expr = \thisrowno{0}, y = r0107:u ] \convtablethree ;
      \addlegendentry{Cycles} ;
      \addlegendentry{Alias} ;
    \end{axis}
  \end{tikzpicture}
\end{figure*}

% Table with detailed performance data, use estimated for O2 because it does not need as many columns (!)
\begin{table*}[t]
  \caption{Most interesting performance counters and correlation with cycle count for optimization O2, estimated cost accounting for constant overhead.}
  \label{tab:convstats}
  \pgfplotstabletypeset[
    font=\small,
    int detect, % Output whole numbers for counter values
    col sep=comma,
    columns={Performance counter, Correlation, 0, 2, 4, 8},
    column type=r,
    columns/Performance counter/.style={
      string type, 
      column type=l,
      postproc cell content/.append code={
        \pgfkeysalso{@cell content=\perfctr{##1}}
      }
    },
    columns/Correlation/.style={
      fixed,
      fixed zerofill,
      precision=2
    },
    every head row/.style={
      output empty row,
      before row={\toprule
        Performance counter & $r$ & 0 & 2 & 4 & 8 \\
        },
      after row=\midrule,
    },
    every last row/.style={after row=\bottomrule}
  ]{bin/conv-default-o2.estimate.csv}
\end{table*}

Figure~\ref{fig:conv-default} illustrates how the convolution kernel behaves for increasing offsets between heap addresses (modulo 4096), clearly indicating a relationship between address aliasing events and cycles executed.
The effects are most distinct on optimization level 2 and 3, where the ratio of cycles to alias events is most significant.
Numbers on the x axis represent the amount of offset, measured in number of \texttt{sizeof(float)} bytes.
An offset of zero is the default behavior for this program when using \texttt{malloc} and moderately large inputs.
Both for optimization level 2 and 3, the default alignment is close to worst case performance.
The differences in cycles executed is also very significant, with about $1.7x$ speedup for O2 and as much as $2x$ speedup for O3.
This phenomenon is only observed for address offsets close to zero, thus we only show the first 32 data points.
If extended to cover the full width of 1024 possible offsets within a 4K segment, graph shows that the performance is uniform everywhere else.

More detailed performance counter data is presented in Table~\ref{tab:convstats}, where we have selected a subset that seem to correlate well with the change to cycle count.
The numbers are for the level 2 optimization case, which was chosen because the behavior was less chaotic and thus more easily represented by only a few numbers.
However, we find the performance data looks fairly similar between the two data sets, with a couple of events that seem to stand out.
\begin{itemize}
    \item A high number of resource stalls for the default alignment, which is reduced substantially together with increasing offsets. 
    \item A high number of cycles with memory loads pending, indicating that the pipeline is stalled waiting for load operations to resolve.
    \item Changes to micro-ops executed for certain ports.
    \item Offcore requests. % todo!
\end{itemize} 
Interestingly, it looks like small address offsets incur a massive increase in operations issued on port 0 for the O2 case. 
On Haswell, this port handles various ALU operations, and branching together with port 6.~\cite[Figure 2.1]{OptimizationManual}
As there is also variations in the number of branch instructions executed, it seems like these counters together show that certain branch micro-ops are re-issued.

It is also worth noticing that most cache related metrics does \emph{not} stand out in this experiment.
The L1 hit rate remains stable across all offsets, and only a neglible number of memory loads actually misses L1.

Overall it seems reasonable to conclude that the resource stalling is causing delayed execution, ultimately generated by false dependencies from address aliasing.
Variations in execution port activity, and in this case increased number of branches executed, is consistent with micro-ops being reissued after the potential conflict is detected.

We do not attempt to pinpoint exactly which assembly instructions might alias, but on a high level we can assume that the processor thinks memory accesses to \texttt{input[i]} potentially conflicts with \texttt{output[i]}.
By manually adjusting the alignment of one of the input buffers, cycle count can be reduced by as much as 50\%.
This is a speedup on top of aggressive compiler optimization, which is not able to account for aliasing effects.

\subsection{Behavior of other Allocators}
As with the contents of environment variables, the exact behavior of dynamic memory allocation can typically not be known ahead of time.
Heap allocation routines such as \texttt{malloc} and \texttt{free} are usually dynamically linked and resolved at runtime, which again represents a potential source of bias.
For the code examples presented earlier, the dynamic linker will resolve \texttt{malloc} to some version of glibc, which is the default on our system.
Variations between different versions of libc, or local configurations on parameters such as thresholds for when to use \texttt{mmap} over \texttt{brk}, could trigger aliasing cases.

\begin{table*}[t]
  \caption{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}
  \label{tab:mallocompare}
  \pgfplotstabletypeset[
    font=\small,
    col sep=comma,
    string type,
    columns={Allocation, 64, 5120, 65536, 1048576, 4194304},
    column type=r,
    columns/Allocation/.style={
      string type, 
      column type=l
    },
    every head row/.style={
      output empty row,
      before row={\toprule
         & 64 B & 5,120 B & 65,536 B & 1,048,576 B & 4,194,304 B \\
        },
      after row=\midrule,
    },
    every odd row/.style={after row=\midrule},
    every last row/.style={after row=\bottomrule}
  ]{bin/malloc-comparison.csv}
\end{table*}

Table \ref{tab:mallocompare} illustrates some differences that can be observed by switching heap allocator\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}.
In addition to glibc, where the heap allocator is called \emph{ptmalloc}, we look at the following alternatives:
\begin{itemize}
  \item Thread-Caching Malloc (tcmalloc) from Google. It claims to be time- and space efficient, and reduces lock contention for multithreaded applications~\cite{TCMalloc}.
  \item jemalloc was originally developed for FreeBSD. It aims to be a scalable general purpose allocator in single or multithreaded environments~\cite{JEMalloc}.
  \item Hoard is another project that focuses extensively on improving performance for multithreaded programs~\cite{Berger:2000:Hoard}.
\end{itemize}
The addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} are observed for different size parameters. 
We do not care about performance, only whether the actual memory addresses returned alias or not.

For smaller allocations, we see that glibc and tcmalloc uses the normal heap area -- returning numerically low addresses.
Interestingly, jemalloc and Hoard appears to never use the normal heap, but allocate to memory mapped areas even for smaller sizes.
We see one example where one allocator yields aliasing buffers while another does not.
Allocating $2 \times 5120$ bytes returns aliasing pointers for jemalloc and Hoard, but not with glibc or tcmalloc.
Given that these results are deterministic (with ASLR disabled), it is not hard to construct a program with significant bias towards one or the other allocator.

%The purpose of this is to illustrate that allocators is yet another source of bias.
Using heap allocated memory, address aliasing is an artifact of the allocator used.
With different versions or configurations of memory allocators, any performance impact from aliasing can qualify as measurement bias.

% Most malloc alternatives focuses heavily on performance in multithreaded environments.
% Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.



\subsection{Ways to Deal with Address Aliasing}
Measurement bias from address aliasing can be significant in real applications, with potentially large performance penalties.
However, with an understanding of the underlying mechanism that causes bias, the effects can to some extent be predicted and accounted for in software.
The most relevant scenario to consider is probably code that can hit the aliasing \texttt{mmap} scenario, where performance impact can be consistently bad over all environments.

\paragraph{Mark buffers with \texttt{restrict}}
In our convolution kernel implementation, the compiler has to account for the fact that input and output pointers might alias, or that the buffers partially overlap.
This limits the extent generated code can keep data in registers without updating the values from memory, as a write to one buffer potentially could invalidate a cached value from the other.
The C99 keyword \emph{restrict} can be used to explicitly tell the compiler that accesses through a pointer does not alias with any other, allowing for more efficient code generation with fewer memory accesses.
\begin{lstlisting}[breaklines=true]
    void conv(int n, const float * restrict input, float * restrict output);
\end{lstlisting}
With this updated function prototype, the number of alias events is reduced by about 10 million on optimization level O2 for the default alignment, with a corresponding improvement in cycle count. % File backing this in results

\paragraph{Use a special purpose allocator}
We found that heap allocators are prone to generate pairwise aliasing buffers, the worst case for functions such as the convolution example.
A potential solution could be to apply some heuristic to randomize addresses more, and in particular not always return the same 12 bit suffix for large allocations.
This goes somewhat agains conventional wisdom that more alignment is better.
The Intel optimization manual actually mentiones this in \emph{User/Source Coding Rule 8}, suggesting that special purpose allocators could be used to avoid aliasing ~\cite{OptimizationManual}.
However, to our knowledge there are no commonly used allocators that specifically tries to mitigate aliasing in heap allocated memory.

\paragraph{Manually adjust memory}
In some cases it might make sense to explicitly control the memory addresses used, for example forcing some fixed relative offset between input and output pointers.
This can be achieved by exploiting the fact the \texttt{mmap} is in fact guaranteed to be placed at a page boundary, and use that directly instead of \texttt{malloc}.
The following approach can be used to make an anonymous memory mapping with offset \texttt{d} bytes away from page alignment.
\begin{lstlisting}[breaklines=true]
    mmap(NULL, (n + d), PROT_READ | PROT_WRITE, MAP_PRIVATE | MAP_ANONYMOUS, -1, 0) + d;
\end{lstlisting}
The same pointer difference must of course by subtracted before unmapping the memory.


\section{Conclusions}
In this paper we have looked at how address aliasing can affect program performance under different memory layouts, and how it can explain certain cases of measurement bias. 
The effect is caused by the way speculative and out-of-order memory operations are handled by the CPU, only considering the last 12 address bits to resolve conflicts between load and store operations.

By analyzing an example originally presented by Mytkowicz et al. \cite{Mytkowicz:2009:WrongData}, we determined that collisions between stack variables and static data resulted in aliasing for certain stack positions.
In this case, the external trigger of bias is variations in environment size, which in turn offset the virtual addresses of stack allocated variables.

In general, any change to virtual memory layout of data can potentially introduce bias effects from address aliasing. 
We present a code example with extreme sensitivity to heap address alignment, with more than 50 \% performance variation between different memory layouts.
We find that typical heap allocators return page-aligned memory by default, which is the worst case with respect to aliasing for many algorithms.
Despite recommendations from Intel, we are not aware of any allocators that specifically addresses this issue. 
We show how manual padding of variables, and alternative alias-free code paths, can be used to avoid aliasing at runtime.
This means that an understanding of 4K address aliasing is sometimes needed to achieve optimal performance. 


\section{Acknowledgements}
This work is forked from a Master's thesis project under the supervision of Anne Cathrine Elster and Rune Erlend Jensen at IDI, NTNU, studying sources of measurment bias for the Intel Ivy Bridge architecture~\cite{MasterThesis}.


\bibliographystyle{plain}
\bibliography{references}

\end{document}
